Type-Safe Modular Parsing

### 2017-01-09 paper/sections/Discussion.tex
Polishing section 4 and 5

\section{Discussion}\label{sec:discussion}

The use of interface-based approach and Object Algebras enables us to do modular parsing. Programmers can easily follow the pattern
captured in the previous section, build modular ASTs and hence have access to multiple dimensions of extensibility. This section
explores more interesting features of this pattern, including overriding existing parsing rules, and framework of language components
for modularity over hierarchies.

@@

\lstinputlisting[linerange=16-22]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=EXTPARSER_TYPEDLAM
On the other hand, the lexer can also be updated, since keywords and delimiters are represented by sets of strings.
One can even ``eliminate'' a production rule in the extension, by actually overriding it with a failure parser.

@@

\subsection{Framework of a Language Component}\label{subsec:framework}

@@

The framework we use for a language, or rather a language component in our case study, consists of three parts: syntax, parsing and operations on the AST. More
specifically:
\begin{itemize}
    \item \textbf{Object Algebra interface:} defined as a trait for the abstract syntax. The type parameters represent multiple sorts of syntax, and the methods are constructs.
\item \textbf{Parser:} implementation of the parser with lexing, using inheritance and Object Algebras, with its scoping trait abstracted over the type parameters.
\item \textbf{Algebra (optional):} some behaviors (operations) on the data structure, such as pretty-printing and so on.
\end{itemize}

@@

\subsection{Fusion of Concepts}\label{subsec:fusion}

\haoyuan{Not sure if this section should be here, or in the related work. I tend to have a big discussion section.}
\huang{These alternatives could be discussed in the discussion
  section, not here. 6.1 could only talks about library code. And the
  titles should be changed as well}\bruno{I think this subsection is
  unnecessary. The code will be presented already at the end of S5, I
  presume. }

In the previous sections we have talked about Packrat Parsing, open recursion and Object Algebras. These components fuse together in our library implementation, for the initial goal, which is to build extensible parsers. We want to argue here that those aspects are totally orthogonal, and hence can have their alternatives. For example, Object Algebras are helpful to build extensible data structures, whereas in functional programming, there has been a large amount of related work on extensible datatypes, including DTC and MRM for the Haskell language. And a better
parsing library would potentially introduce a lot of fancy features and parser combinators, run with high efficiency, or even support indirect left-recursion.
\haoyuan{Open recursion could also have alternatives?} For now, our implementation integrates the existing ones, and has turned out to be practical in various applications.

>>>>

\section{More Features}

The use of inheritance-based approach and Object Algebras enables us to build modular parsers, which are able to evolve with syntax together. This section explores more interesting features, including overriding existing parsing rules, language components for abstracting language features, and alternative techniques under the whole framework.

@@

\lstinputlisting[linerange=16-22]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=EXTPARSER_TYPEDLAM
Futhermore, the lexer can also be updated, since keywords and delimiters are represented by sets of strings.
One can even ``eliminate'' a production rule in the extension, by actually overriding it with a failure parser.

@@

\subsection{Language Component}

@@

The framework we use for a language, or rather a language component in our case study, consists of three parts: syntax, parsing and operations on the AST. More
specifically:
\begin{itemize}
    \item \textbf{Object Algebra interface:} defined as a trait for the abstract syntax. The type parameters represent multiple sorts of syntax, and the methods are constructs.
\item \textbf{Parser:} implementation of the parser with lexing, using inheritance and Object Algebras, with its scoping trait abstracted over the type parameters.
\item \textbf{Algebras (optional):} concrete operations on ASTs, such as pretty-printing.
\end{itemize}

@@

\subsection{Alternative Techniques}
\huang{to rewrite}
In the previous sections we have talked about Packrat Parsing, open recursion and Object Algebras. These components fuse together in our library implementation, for the initial goal, which is to build extensible parsers. We want to argue here that those aspects are totally orthogonal, and hence can have their alternatives. For example, Object Algebras are helpful to build extensible data structures, whereas in functional programming, there has been a large amount of related work on extensible datatypes, including DTC and MRM for the Haskell language. And a better
parsing library would potentially introduce a lot of fancy features and parser combinators, run with high efficiency, or even support indirect left-recursion.
\haoyuan{Open recursion could also have alternatives?} For now, our implementation integrates the existing ones, and has turned out to be practical in various applications.

### 2017-01-09 paper/sections/ObjectAlgebras.tex
Polishing section 4 and 5

\section{Extensible ASTs with Object Algebras}\label{sec:algebrasandparsing}

@@

Now it comes to Extension 2 in Section~\ref{subsec:parsingproblem}. In the previous code, we have presented pretty-printing as
an example of operations on ASTs. Suppose we also want to collect free variables on the expression language which models
literals, additions and variables. Based on the traditional OO class hierarchy, one attempt would be extending \inlinecode{Expr} with the new
operation directly:

@@

It refers to one Scala solution to the Expression Problem.\huang{what's the solution? is it successful?} But then programmers have to define new classes for literals, additions and variables
by extending the old ones together with \inlinecode{FreeVarsExpr}. The critical issue is that, the existing parsing code only generates objects of the old
types, again it is a modularity issue, since it is a bad idea to modify existing code again and again with extensions. Other solutions including modelling constructors as functions introduce unnecessary complexity.

@@

\subsection{Object Algebras}\label{subsec:objectalgebras}
\huang{to polish}Object Algebras captures a design pattern to address the Expression Problem nicely,
achieving two dimensions of extensibility (data variants and operations) in a modular and type-safe way.
Because of this, the definition of data structures is separated from behaviors on them, and future extensions
to both sides no longer require existing code to be modified, supporting separate compilation.

@@

\lstinputlisting[linerange=32-34]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EXTPRINT
We observe that only the new case is implemented for pretty-printing, and the others have been inherited.
It offers programmers great code reuse, in the sense that existing cases can also be
overridden in \inlinecode{VarExprPrint}.

To create an expression of \inlinecode{VarExprAlg}, a generic method is defined as follows:

\lstinputlisting[linerange=41-41]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_MAKEEXP
Such a method implicitly represents the expression \inlinecode{"1 + x"}. The code

\begin{lstlisting}
makeExp(new VarExprPrint {})
\end{lstlisting}
results in \inlinecode{"(1 + x)"}, as the result of pretty-printing. Now we observe that an ``object'' of expression actually
has the function type \inlinecode{VarExprAlg[E] =>} \inlinecode{E} for generic \inlinecode{E}. At this point, one may argue
that it is illusory compared to traditional objects, and since we are struggling against the parsing problem, if a parser
produces structures as functions, it could be hardly be further processed like objects, especially, in the design of a compiler
there will be a lot of desugarings (or transformations). However, we argue that this is not an issue, because functions can be used
in a very clever way. For example, if one wants to transform a \inlinecode{VarExprAlg} expression to an \inlinecode{ExprAlg} expression by
replacing some variables with values based on a variable environment, a transformation algebra can be implemented as follows:\haoyuan{mention in discussion that shy deals with boilerplate.}

@@

Just as shown above, \inlinecode{ExprAlg} is defined for a small language in an Object-Algebra style. Furthermore,
A parser can be defined for this whole language at once. Since Object Algebras represent ``objects'' implicitly as functions like \inlinecode{ExprAlg[E] =>} \inlinecode{E} for generic \inlinecode{E}, such a parser should have type \inlinecode{ExprAlg[E] =>} \inlinecode{Parser[E]}, where \inlinecode{E} can be abstracted by the enclosing type.

\lstinputlisting[linerange=8-16]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
Such a parser works fine, yet it does not appear to be extensible. For example, we have demonstrated in Section~\ref{subsec:inheritance-approach} that method overriding can update \inlinecode{pExpr} for an extended syntax, but now \inlinecode{pExpr} will have a different type, like \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]} where the difference appears in the argument. The difference of types prohibits method overriding, however, this problem refers to \textit{binary methods}. \haoyuan{Shall we mention this?}

One solution to this problem is taking out the argument as a single field, hence it supports overriding as well. Therefore we refactor the parsing code as follows:
\lstinputlisting[linerange=23-33]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER
This is precisely the pattern that we advocate for modular parsing. Note that \inlinecode{pE} is the one that is used recursively, exported by the trait and considered to be overridden from inheritance. The reason why we use an extra and seemingly redundant \inlinecode{pExpr}, is that \inlinecode{pE} can no longer be referred to by the \inlinecode{super} keyword, as a limitation of Scala for generics, hence we can directly invoke \inlinecode{pExpr} as the specific \inlinecode{pE} in \inlinecode{ExprOAParser}. And we still want code reuse for the composition of \inlinecode{pLit} and \inlinecode{pAdd}. Now it is worth trying on this pattern with an extension, like \inlinecode{VarExprAlg} which we have been using:
\lstinputlisting[linerange=37-48]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER
In the client code shown above, an instance of algebra is fed to \inlinecode{VarExprOAParser} for pretty-printing expressions. Then the trait exports \inlinecode{pE}, parses the input and applies the algebra. The result shows that such a pattern is a nice solution to modular parsing. At this point, one may argue that with a different operation like evaluation, the redundant code is required to obtain the parser, and it results in parsing the same input twice. This is generally solved in the original Object Algebras paper [] by composing two algebras into one in advance.

\subsection{Parsing with Multiple Syntax}\label{subsec:differentsyntax}

As illustrated above, one intuition of using Object Algebras is to separate data structures from behaviors, so as to modularize the design, and both parts are even more modular and reusable. We can simply add new variants corresponding to the new cases in the grammar, and realize different operations on structures, with both code reuse and separate compilation supported.

Another advantage is that Object Algebras benefit us from using generics, where we can abstract multiple syntax in the type parameters, and define less classes, simplifying code comparatively. Below we further extend the language with types, by introducing an extra type parameter in the algebra interface. The extended grammar is:\\

@@

We write code for the extended AST, pretty-printing operation and the extended parser:

@@

\lstinputlisting[linerange=63-65]{../Scala/Parser/src/PaperCode/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX_CLIENT
Pretty-printing the parsed results just works as expected.

>>>>

\section{Full Extensibility with Object Algebras}\label{sec:algebrasandparsing}

@@

Now it comes to Extension 2 in Section~\ref{subsec:parsingproblem}. In the previous code, we have pretty-printing as
an example of operations on ASTs. Suppose we also want to collect free variables on the expression language which models
literals, additions and variables. Based on the traditional OO class hierarchy, one attempt would be extending \inlinecode{Expr} with the new
operation directly:

@@

It refers to one Scala solution to the Expression Problem.\huang{what's the solution? does it have the later "critical issue"? if so, why is it a "solution"} But then programmers have to define new classes for literals, additions and variables
by extending the old ones together with \inlinecode{FreeVarsExpr}. The critical issue is that, the existing parsing code only generates objects of the old
types, again it is a modularity issue, since it is a bad idea to modify existing code again and again with extensions. Other solutions including modelling constructors as functions introduce unnecessary complexity.

@@

\subsection{Object Algebras}\label{subsec:objectalgebras}
Object Algebras captures a design pattern to address the Expression Problem nicely, achieving two dimensions of extensibility (data variants and operations) in a modular and type-safe way. The definition of data structures is separated from their behaviors, and future extensions on both dimensions no longer require existing code to be modified, supporting separate compilation.

@@

\lstinputlisting[linerange=32-34]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EXTPRINT
An observation is that only the new case is implemented for pretty-printing, and the others have been inherited.
It offers programmers great code reuse, in the sense that existing cases can also be
overridden in \inlinecode{VarExprPrint}.

To create an expression representing \inlinecode{1 + x}, a generic method is defined as follows:

\lstinputlisting[linerange=41-41]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_MAKEEXP

The code below results in \inlinecode{"(1 + x)"}, as the result of pretty-printing that expression.

\begin{lstlisting}
makeExp(new VarExprPrint {})
\end{lstlisting}

Now we observe that an ``object'' of expression actually
has the function type \inlinecode{VarExprAlg[E] =>} \inlinecode{E} for generic \inlinecode{E}. Comparing with traditional objects, one may have concern that if a parser produces function structures, it could be hard to further process them. Especially, in the design of a compiler
there will be a lot of desugarings (or transformations). However, we argue that this is not an issue, because functions can be used
in a very clever way. For example, if one wants to transform a \inlinecode{VarExprAlg} expression to an \inlinecode{ExprAlg} expression by
replacing some variables with values based on a variable environment, a transformation algebra can be implemented as follows:\haoyuan{mention in discussion that shy deals with boilerplate.}

@@

Just as shown above, \inlinecode{ExprAlg} is defined for a small language using Object Algebras. Since Object Algebras represent ``objects'' implicitly as functions like \inlinecode{ExprAlg[E] =>} \inlinecode{E} for abstract \inlinecode{E}, we first try to build the corresponding parser by type \inlinecode{ExprAlg[E] =>} \inlinecode{Parser[E]}.

\lstinputlisting[linerange=8-16]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
Such a parser works fine, yet it does not appear to be extensible. For example, we have demonstrated in Section~\ref{subsec:overriding} that method overriding is essential to update \inlinecode{pExpr} for an extended syntax. However, now \inlinecode{pExpr} has type \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]}, where the type of Object Algebra interface appears in \emph{contravariant} position. This difference prohibits method overriding, because Object Algebra interfaces extended from \inlinecode{VarExprAlg} are subtypes of it, but the types of their parsers are supertypes of \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]}.

This problem refers to \textit{binary methods}. \haoyuan{Shall we mention this?}\huang{???}

One solution to this problem is taking out the argument as a single field, hence it supports overriding as well. We rewrite the parsing code as follows:

\lstinputlisting[linerange=23-33]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER
This is precisely the pattern that we advocate for modular parsing. Note that \inlinecode{pE} is the one that is used for recursive calls, exported by the trait and considered to be overridden during inheritance. The reason why we use an extra and seemingly redundant \inlinecode{pExpr}, is a subtle issue caused by Scala language and its parser combinator library. There is a restriction of \inlinecode{super} keyword in Scala that \inlinecode{super} can only use fields defined by keyword \inlinecode{def}, can not use those defined by \inlinecode{val}, while the parser combinator library suggests using \inlinecode{val} to define parsers, especially for left-recursive ones. Our workaround is that we use different synonyms for \inlinecode{pE} in different traits, so that we can directly distinguish them by names without using \inlinecode{super}.

Now let's try this pattern with the extension of varibles, which we have been using:
\lstinputlisting[linerange=37-48]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER
In the client code shown above, an instance of algebra is fed to \inlinecode{VarExprOAParser} for pretty-printing expressions. Then the trait exports \inlinecode{pE}, parses the input and applies the algebra. The result shows that such a pattern is a nice solution to modular parsing. At this point, one may argue that with a different operation like evaluation, the redundant code is required to obtain the parser, and it results in parsing the same input twice. However, that is generally solved~\cite{Oliveira2012} \huang{it was "original object algebra paper", i added the citation here, is it the correct one? i'll check later} by composing two algebras into one in advance.

\subsection{Parsing Multi-sort Syntax}\label{subsec:differentsyntax}

\huang{to polish}As illustrated above, one intuition of using Object Algebras is to separate data structures from behaviors, so as to modularize the design, and both parts are even more modular and reusable. We can simply add new variants corresponding to the new cases in the grammar, and realize different operations on structures, with both code reuse and separate compilation supported.

Another advantage of using Object Algebras is that we benefit from using generics. If the syntax contains multiple sorts, we can distinguish them by different type parameters. Thus fewer classes are defined and code is simplified. As an example, we extend the language of literals, additions and variables by introducing types as a new sort. Now it has two sorts in the syntax, types and expressions. We certainly do not want to mix them together.\\

@@

When extending the Object Algebra interface, pretty-printing operation and parser, we use two type parameters \inlinecode{E} and \inlinecode{T} for expressions and types. They guarantee that invalid terms such as \inlinecode{int + int} will be rejected by the parser.

@@

\lstinputlisting[linerange=63-65]{../Scala/Parser/src/PaperCode/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX_CLIENT
It works just as expected to parse and pretty-print the result.
